---
---
title: "Tidy Data for fluorescent marker"
subtitle: "Exicitation and Emission Spectra for Fluorescently tagged SMOB Reductase"
author: Tymesha Bovell
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

This data contains excitation and emission fluorscence spectra for the mScarlett fluorescent protein fused to my SMOB reductase enzyme. This datset includes wavelength dependent fluorescence intensity measurements
## About the data

The dataset used here is `Emission.csv` and `Excitation.csv` 

-   Date downloaded: Febuary 26th, 2026
-   Download source: 
You will explore the nitrogen to phosphate ratio, which indicates the health of the ocean ecosystem. <https://beal-agulhas.earth.miami.edu/teaching/courses/lecture-five/index.html>

## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(here)
library(readr)
```

# Read in the data

Quarto runs R in the same way as a script: objects only exist after the chunks that create them have been run.

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
biochem_em <- read_csv(here::here("data/raw",  "Emission.csv"))
biochem_ex <- read_csv(here::here("data/raw", "Excitation.csv"))



#(Optional) Quick peek at the data:

# Uncomment if you want more inspection output
 names(biochem_em)
 names(biochem_ex)
# str(biochem_em)
 # str(biochem_ex)
# head(biochem_em)
# head(biochem_ex)
# summary(biochem_em)
# summary(biochem_ex)
```
