---
title: "Working with Spatial Data in R"
author: "Tymesha Bovell"
editor: visual
---

# Introduction

This project explores spatial data analysis and visualization in R using the `sf` package and associated geospatial tools. The analysis focuses on Alaska regional boundaries, Alaska river systems, and Alaska community population datasets to demonstrate how spatial joins, coordinate reference systems (CRS), and mapping techniques can be used to connect demographic information with geographic regions.

The regional boundary dataset comes from the State of Alaska’s Salmon and People (SASAP) project developed by Jared Kibele and Jeanette Clark (2018). The SASAP regions were updated to better reflect HUC 8 watershed boundaries while correcting topological issues such as overlaps and gaps. These polygon shapefiles provide the primary regional framework used throughout the analysis.

Community population and household language data were derived from census and American Community Survey datasets compiled by Jeanette Clark, Sharis Ochs, Derek Strong, and the National Historic Geographic Information System (NHGIS). The data include community-level population information and geographic coordinates that allow the communities to be spatially mapped and joined to Alaska regions.

Additional context was provided through a rivers shapefile representing rivers grouped by SASAP region. This dataset was derived from the USGS Hydrographic Geodatabase and the Riverscape Analysis Project and was simplified for visualization purposes.

Using these datasets, this project demonstrates how spatial data can be transformed, joined, summarized, and visualized using both static and interactive mapping workflows in R.

# Data:

**Source 1:**\
Kibele, J., & Clark, J. (2018). *State of Alaska’s Salmon and People Regional Boundaries*. Knowledge Network for Biocomplexity. https://doi.org/10.5063/F1125QWP

Date of retrival:04/30/2026

**Source 2:**\
Clark, J., Ochs, S., Strong, D., & National Historic Geographic Information System. (2018). *Languages used in Alaskan households, 1990–2015*. Knowledge Network for Biocomplexity. https://doi.org/10.5063/F11G0JHX

Date of retrival:04/30/2026

**Source 3:**\
Kibele, J., & Clark, J. (2019). *Rivers of Alaska grouped by SASAP region, 2018*. Knowledge Network for Biocomplexity. https://doi.org/10.5063/F1PZ573F

Date of retrival:04/30/2026

```{r}
#Loading libraries and packages
library(readr)
library(here)
library(sf)
library(ggplot2)
library(leaflet)
library(scales)
library(ggspatial)
library(dplyr)
library(prettymapr)

```

```{r}
# read in shapefile using read_sf()
ak_rgns_sf <- read_sf(here("data/raw/Alaska/ak_regions_simp.shp"))

#quick plot
#plot(ak_rgns_sf)

#examine its class using
#(ak_rgns_sf)

#view crs
st_crs(ak_rgns_sf)
```

## CRS for datum and Projection

```{r}
#transformgeospatial data from one coordinate system to another.
ak_rgns_3338_sf <- ak_rgns_sf %>%
    st_transform(crs = 3338)

st_crs(ak_rgns_3338_sf)

plot(ak_rgns_3338_sf)

```

## sf & the Tidyverse

```{r}
# returns the names of all the columns in dataset
colnames

ak_rgns_3338_sf %>%
    select(region)
#variable to filter specific region
unique(ak_rgns_3338_sf$region)
ak_rgns_3338_sf %>%
    filter(region == "Southeast")
```

## Spatial Joins for different datasets

```{r}
# read in population data
pop_df <- read_csv(here("data/raw/Alaska/alaska_population.csv"))

pop_4326_sf <- st_as_sf(pop_df,
                        coords = c('lng', 'lat'),
                        crs = 4326,
                        remove = F)

head(pop_4326_sf)

#Join population data with Alaska regions data with matched crs
pop_3338_sf <- st_transform(pop_4326_sf, 
                            crs = 3338)
pop_joined_sf <- st_join(pop_3338_sf, 
                         ak_rgns_3338_sf, 
                         join = st_within)

head(pop_joined_sf)

#Compute total population by region
pop_rgn_df <- pop_joined_sf %>%
    st_drop_geometry() %>%
    group_by(region) %>%
    summarize(total_pop = sum(population))

head(pop_rgn_df)

pop_rgn_3338_sf <- left_join(ak_rgns_3338_sf, 
                             pop_rgn_df, 
                             by = "region")

# plot to check
plot(pop_rgn_3338_sf["total_pop"])

# calculate the population by Alaska management area

pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop))

plot(pop_mgmt_3338_sf["total_pop"])

# Uncombine geometries

pop_mgmt_3338_sf <- pop_rgn_3338_sf %>%
    group_by(mgmt_area) %>%
    summarize(total_pop = sum(total_pop), do_union = F)

plot(pop_mgmt_3338_sf["total_pop"])

#Save the spatial object to a new file using 
write_sf(pop_rgn_3338_sf, here("data/ak_regions_population.shp"))

#ggplot visualization
ggplot(pop_rgn_3338_sf) +
    geom_sf(aes(fill = total_pop)) +
    labs(fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()

#visual of rivers in Alaska and location 
rivers_3338_sf <- read_sf(here("data/raw/Alaska/ak_rivers_simp.shp"))
st_crs(rivers_3338_sf)

ggplot() +
    geom_sf(data = pop_rgn_3338_sf, 
            aes(fill = total_pop)) +
    geom_sf(data = pop_3338_sf, 
            size = 0.5) +
    geom_sf(data = rivers_3338_sf,
            aes(linewidth = StrOrder)) +
    scale_linewidth(range = c(0.05, 0.5),
                    guide = "none") +
    labs(title = "Total Population by Alaska Region",
         fill = "Total Population") +
    scale_fill_continuous(low = "khaki",
                          high =  "firebrick",
                          labels = comma) +
    theme_bw()


# Base maps into spatial maps using ggspatial
pop_3857_sf <- st_transform(pop_3338_sf, 
                         crs = 3857)
ggplot(data = pop_3857_sf) +
    ggspatial::annotation_map_tile(type = "osm", zoom = 4, progress = 'none') + 
    # higher zoom values are more detailed 
    geom_sf(aes(color = population),
            fill = NA) +
    scale_color_continuous(low = "darkkhaki",
                           high =  "firebrick",
                           labels = comma)

#Visualize sf objects with leaflet for Alaska Albers
epsg3338 <- leaflet::leafletCRS(
    crsClass = "L.Proj.CRS",
    code = "EPSG:3338",
    proj4def =  "+proj=aea +lat_1=55 +lat_2=65 +lat_0=50 +lon_0=-154 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs",
    resolutions = 2 ^ (16:7)
)

#unprojected coordinate system for leaflet
pop_rgn_4326_sf <- pop_rgn_3338_sf %>% 
    st_transform(crs = 4326)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(data = pop_rgn_4326_sf,
                fillColor = "gray",
                weight = 1)

m

#adding labels, legends and color scale
pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1,
        label = ~ region
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

m

# Display individual communities with respective labels for population

pal <- colorNumeric(palette = "Reds", domain = pop_rgn_4326_sf$total_pop)

m <- leaflet(options = leafletOptions(crs = epsg3338)) %>%
    addPolygons(
        data = pop_rgn_4326_sf,
        fillColor = ~ pal(total_pop),
        weight = 1,
        color = "black",
        fillOpacity = 1
    ) %>%
    addCircleMarkers(
        data = pop_4326_sf,
        lat = ~ lat,
        lng = ~ lng,
        radius = ~ log(population / 500),
        # arbitrary scaling
        fillColor = "gray",
        fillOpacity = 1,
        weight = 0.25,
        color = "black",
        label = ~ paste0(pop_4326_sf$city, ", population ", comma(pop_4326_sf$population))
    ) %>%
    addLegend(
        position = "bottomleft",
        pal = pal,
        values = range(pop_rgn_4326_sf$total_pop),
        title = "Total Population"
    )

m
```
# Discussion:


This project demonstrated how spatial data can be transformed, joined, summarized, and visualized in R using the `sf` package and tidyverse workflows. The analysis combined Alaska regional boundary polygons, Alaska community population point data, and river shapefiles to create both static and interactive geospatial visualizations.

One of the most important parts of the analysis was transforming both the Alaska region polygons and the population point data into Alaska Albers (EPSG:3338). This decision was necessary because the spatial join would not work correctly when the datasets used different coordinate reference systems. Using Alaska Albers also improved the appearance of the Alaska map compared to WGS84 and provided a projection better suited for the study area.

The analysis used `st_within()` to assign Alaska communities to their respective regions and summarize population by region. Spatial data were then visualized using `ggplot2`, `ggspatial`, and `leaflet` to display population distributions and regional patterns across Alaska.

Overall, this project provided practical experience working with coordinate reference systems, spatial joins, geospatial visualization, and spatial data manipulation in R. The workflow demonstrated how multiple spatial datasets can be integrated to create regional population analyses and geographic visualizations.
